---
title: "What Hath Von Neumann Wrought?"
author: "M. Edward (Ed) Borasky"
date: "June 18, 2016"
output: ioslides_presentation
---
# Motivation

## Standing on the shoulders of giants

## Born with Curiosity

## This was done with a slide rule

## Archive.org and bitsavers

## The world of Grace Hopper

## The Harvard Mark I

## What was your first computer?

## Did you learn assembler?

## Did you learn machine code?

# John von Neumann

## In Europe

## The Manhattan Project

## Game Theory

## Self-reproducing systems / cellular automata

## Applied Mathematics

## Bureaucrat

## Death

# The IAS Machines

## State of scientific computing

## Preliminary discussions

## Planning and coding

## Hardware
* MANIAC
* JOHNNIAC
* ILLIAC and copies

## Commercial variants
* IBM 701
* Univac 1100

# The architecture

## Arithmetic unit

```{r echo = FALSE}
library(DiagrammeR)
registers <- create_nodes(nodes = c(
  "Memory", "Accumulator", "Adder", "Quotient", "R3", "Complement Gate"),
  shape = "rectangle",
  width = 2)
arith_unit <- create_graph(registers)
arith_unit <- add_edge(arith_unit, "Accumulator", "Memory")
arith_unit <- add_edge(arith_unit, "Accumulator", "Adder")
arith_unit <- add_edge(arith_unit, "Adder", "Accumulator")
arith_unit <- add_edge(arith_unit, "Memory", "Quotient")
arith_unit <- add_edge(arith_unit, "Memory", "R3")
arith_unit <- add_edge(arith_unit, "Quotient", "R3")
arith_unit <- add_edge(arith_unit, "R3", "Complement Gate")
arith_unit <- add_edge(arith_unit, "Complement Gate", "Adder")
render_graph(arith_unit)
```

## The accumulator (A) and quotient register (Q) coupled

```{r echo = FALSE}
library(DiagrammeR)
bits <- create_nodes(nodes = c(
  "a0", "a1", "a-middle", "a38", "a39", 
  "q0", "q1", "q-middle", "q38", "q39"),
  shape = "rectangle")
aq <- create_graph(bits, graph_attrs = "rankdir = TB")
render_graph(aq)
```

## Right Shifting

```{r echo = FALSE}
bits <- create_nodes(nodes = c(
  "a0", "a1", "a-middle", "a38", "a39", 
  "q0", "q1", "q-middle", "q38", "q39"),
  shape = "rectangle")
right_shift <- create_graph(bits, graph_attrs = "rankdir = LR")
right_shift <- add_edge(right_shift, "a0", "a0")
right_shift <- add_edge(right_shift, "a0", "a1")
right_shift <- add_edge(right_shift, "a1", "a2")
right_shift <- add_edge(right_shift, "a2", "a-middle")
right_shift <- add_edge(right_shift, "a-middle", "a38")
right_shift <- add_edge(right_shift, "a38", "a39")
right_shift <- add_edge(right_shift, "a39", "q1")
right_shift <- add_edge(right_shift, "q1", "q2")
right_shift <- add_edge(right_shift, "q2", "q-middle")
right_shift <- add_edge(right_shift, "q-middle", "q38")
render_graph(right_shift)
```

## Left Shifting

## Adding

## Multiplying

## Dividing

## Control

## Input / Output

# The software engineering process

## Algorithms
* Calculus / differential equations
* Statistics / Monte Carlo
* Linear algebra / linear equations / eigensystems

## Memory layout

## Assembler / loader

## Run

## Debugging / post-mortem

# The emulator

## Existing emulators
* simh
* SILLIAC
* misc

## Why build a new one?

