---
title: "What Hath Von Neumann Wrought?"
author: "M. Edward (Ed) Borasky"
date: "June 18, 2016"
output: ioslides_presentation
---
# Motivation

## Standing on the shoulders of giants

## Born with Curiosity

## This was done with a slide rule

## Archive.org and bitsavers

## The world of Grace Hopper

## The Harvard Mark I

## What was your first computer?

## Did you learn assembler?

## Did you learn machine code?

# John von Neumann

## In Europe

## The Manhattan Project

## Game Theory

## Self-reproducing systems / cellular automata

## Applied Mathematics

## Bureaucrat

## Death

# The IAS Machines

## State of scientific computing

## Preliminary discussions

## Planning and coding

## Hardware
* MANIAC
* JOHNNIAC
* ILLIAC and copies

## Commercial variants
* IBM 701
* Univac 1100

# The architecture

## Arithmetic unit

```{r echo = FALSE}
DiagrammeR::grViz("arith_unit.gv")
```

## The accumulator (A) and quotient register (Q) coupled

```{r echo = FALSE}
DiagrammeR::grViz("aq.gv")
```

## Left Shifting = multiply AQ by two
1. Shift a0 -> bit bucket
2. Shift a1:a39 -> a0:a38
3. Leave q0 alone!
4. Shift q1 -> a39
5. Shift q2:q39 -> q1:q38
6. Zero -> q39

## Right Shifting = divide AQ by two
1. Shift q39 -> bit bucket
2. Shift q1:q38 -> q2:q39
3. Leave q0 alone!
4. Shift a39 -> q1
5. Shift a0:a38 -> a1:a39
6. Keep original a0!

## Adding
1. Augend
    * Clear A, or
    * Hold A, or
    * Set A = 1/2
2. Addend
    * Memory word or Q to R3, then
3. Complement gate
    * Pass straight through, or
    * Negate, or
    * Absolute value
3. Form sum in A

## Multiplication: initialization
1. Multiplier is in Q
2. Multiplicand is in R3
3. Accumulator
    * Clear A, or
    * Hold A, or
    * Set A = 1/2
4. Complement gate
    * Pass straight through, or
    * Negate, or
    * Absolute value

## Multiplication: iteration
39 steps of

* A <- A + Complement gate
* Shift AQ right one bit

## Dividing
1. Double-length dividend in AQ
2. Divisor is in R3
3. 39 steps of
    * subtract R3 from A in the adder
    * if the result is less than zero, set q39 to 0
    * if the result is greater than or equal to zero, set q39 to 1 and copy the difference to A
    * shift AQ left one place, copying q1 into q0
4. Set q39 to 1 for rounding

## Control

## Input / Output

# The software engineering process

## Algorithms
* Calculus / differential equations
* Statistics / Monte Carlo
* Linear algebra / linear equations / eigensystems

## Memory layout

## Assembler / loader

## Run

## Debugging / post-mortem

# The emulator

## Existing emulators
* simh
* SILLIAC
* misc

## Why build a new one?

